\large How resource abundance and stochasticity affect organisms' range sizes

### 2024-05-30 writing/appendix-A-concepts-and-definitions.Rmd
updated writing

---
title: \large How resource abundance and stochasticity affect animals' space-use requirements
subtitle: "Appendix A: Concepts and definitions"
# author:
#   - name: Stefano Mezzini
#     email: stefano.mezzini@ubc.ca
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure A\!\!} # for "Figure Ax".
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize} # smaller caption text
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    number_sections: false
    toc: false
    # latex_engine: xelatex # for sigma characters in figure
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua' # for author affiliations
      - '--lua-filter=author-info-blocks.lua' # for author affiliations
---

@@

```{r ex, eval=TRUE, echo=FALSE, fig.height=5, fig.cap="Examples of Gamma distributions with different means ($\\mu$) and variances ($\\sigma^2$).", fig.pos='!h', message=FALSE, warning=FALSE, out.width = '75%'}
library('dplyr') # for data wrangling
library('tidyr') # for data wrangling
library('ggplot2') # for fancy plots

@@

expand_grid(x = tibble(x = seq(0, 10, length.out = 1000)),
            mean = c(1, 5),
            var = c(0.5, 5)) %>%
  unnest(x) %>%
  mutate(dens = dgamma2(x = x, mu = mean, sigma2 = var)) %>%
  ggplot(aes(x, dens)) +
  coord_cartesian(xlim = c(-0.5, 10.5), ylim = c(0, 0.8)) +
  # using '\U03BC' because mu results in \propto, for some reason
  facet_wrap(. ~ mean + var,
             labeller = label_bquote('\U03BC' == .(mean)~','~sigma^2 == .(var)))+
  geom_line(na.rm = TRUE) +
  geom_area(alpha = 0.3, na.rm = TRUE) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(x = 'r', y = 'Density') +
  theme_bw()
```

>>>>

---
title: \large How resource abundance and stochasticity affect organisms' range sizes
subtitle: "Appendix A: Concepts and definitions"
# author:
#   - name: Stefano Mezzini
#     email: stefano.mezzini@ubc.ca
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
# csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure A\!\!} # for "Figure Ax".
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize} # smaller caption text
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    number_sections: false
    toc: false
    # latex_engine: xelatex # for sigma characters in figure
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua' # for author affiliations
      - '--lua-filter=author-info-blocks.lua' # for author affiliations
---

@@

```{r ex-gamma, eval=TRUE, echo=FALSE, fig.height=5, fig.cap="Examples of Gamma distributions with different means ($\\mu$) and variances ($\\sigma^2$).", fig.pos='!h', message=FALSE, warning=FALSE, out.width = '75%'}
library('dplyr') # for data wrangling
library('tidyr') # for data wrangling
library('ggplot2') # for fancy plots

@@

expand_grid(x = tibble(x = seq(0, 10, length.out = 1000)),
            mean = c(1, 5),
            var = c(0.5, 5)) %>%
  unnest(x) %>%
  mutate(dens = dgamma2(x = x, mu = mean, sigma2 = var)) %>%
  ggplot(aes(x, dens)) +
  coord_cartesian(xlim = c(-0, 10), ylim = c(0, 0.8)) +
  # using '\U03BC' because mu results in \propto, for some reason
  facet_wrap(. ~ mean + var,
             labeller = label_bquote('\U03BC' == .(mean)~','~sigma^2 == .(var)))+
  geom_line(na.rm = TRUE) +
  geom_area(alpha = 0.3, na.rm = TRUE) +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 2.5, 5, 7.5)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = 'R', y = 'Density') +
  theme_bw() +
  theme(panel.grid = element_blank())
```

```{r ex-beta, eval=TRUE, echo=FALSE, fig.height=5, fig.cap="Examples of Beta distributions with different means ($\\mu$) and variances ($\\sigma^2$).", fig.pos='!h', message=FALSE, warning=FALSE, out.width = '75%'}
# to help visualize the consequences of different E(R) and V(R) if R ~ Beta
dbeta2 <- function(x, mu, sigma2, lower.tail = FALSE,
                   eps = .Machine$double.eps) {
  phi <- sigma2 / (mu * (1 - mu)) # scale parameter

  dbeta(x = x,
        shape1 = mu * (1 / phi - 1),
        shape2 = (1 - mu) * (1 / phi - 1))
}

expand_grid(x = tibble(x = seq(0, 1, length.out = 1000)),
            mean = c(0.5, 0.75),
            var = c(0.002, 0.05)) %>%
  unnest(x) %>%
  mutate(dens = dbeta2(x = x, mu = mean, sigma2 = var)) %>%
  ggplot(aes(x, dens)) +
  coord_cartesian(ylim = c(0, NA)) +
  # using '\U03BC' because mu results in \propto, for some reason
  facet_wrap(. ~ mean + var,
             labeller = label_bquote('\U03BC' == .(mean)~','~sigma^2 == .(var)))+
  geom_line(na.rm = TRUE) +
  geom_area(alpha = 0.3, na.rm = TRUE) +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.25, 0.5, 0.75, 1),
                     labels = c('0', '0.25', '0.5', '0.75', '1')) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = 'R', y = 'Density') +
  theme_bw() +
  theme(panel.grid = element_blank())
```

### 2024-05-30 writing/appendix-B-simulations.Rmd
updated writing

---
title: \large How resource abundance and stochasticity affect animals' space-use requirements
subtitle: "Appendix B: Simulations"
# author:
#   - name: Stefano Mezzini
#     email: stefano.mezzini@ubc.ca
#     institute: [biol, braes]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [biol, braes, cmps]
# institute:
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
header-includes:
    - \renewcommand{\figurename}{Figure B\!\!} # for "Figure Sx.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize} # smaller caption text
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    toc: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua' # for author affiliations
      - '--lua-filter=author-info-blocks.lua' # for author affiliations
---

@@

```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' space-use requirements were simulated. (a.) Tracks were simulated using an Integrated Ornstein-Uhlenbeck model (IOU model, an infinitely diffusive and continous-velocity movement model), starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square). (b.) Each time the track crossed into a new cell (green dots), the animal collected a random amount of resources that followed a Gamma distribution with common mean $\\mu (t)$ and variance $\\sigma^2 (t)$. (c.) Each time the animal collected more resources, its satiety (purple) increased. Once the animal collected sufficient resources, the animal stopped exploring (i.e., the track was truncated) and was allowed to return to $\\langle0, 0 \\rangle$ over the same amount of time needed to reach satiety. (D) The process was repeated 200 times (13 tracks pictured in this panel). (E) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h', cache=FALSE}
knitr::include_graphics('../figures/simulation-example.png', rel_path = FALSE)
```

@@

**Exploratory plot B\@ref(fig:dt-sensitivity)a.** The amount of time between encounters ranged from 1 second (the minimum sampling interval) to 24 minutes and 10 seconds. Approximately 93% of the encounters (500/536, excluding the first 3 events of each track) occurred with 30 or more seconds between events.

**Exploratory plot B\@ref(fig:dt-sensitivity)b.** Halving the sampling interval had little to no effect on the total number of encounters for $\Delta t \lessapprox 60 \,s.$

**Exploratory plot B\@ref(fig:dt-sensitivity)c.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks but large enough to avoid excessive amounts of data and an inflated amount of encounters when an animal was near cell boundaries.

**Exploratory plot B\@ref(fig:dt-sensitivity)d.** The three tracks used for these exploratory plots are sufficiently different that we considered them to be a representative sample of movement tracks simulated by the OUF model. All three tracks have a reasonable amount of both tortuous and directed movement.

@@

@@

In this script, we check how many tracks are necessary to produce stable and accurate estimates of the space-use requirements. We do so by estimating the home ranges of animals for varying numbers of track in the best-case scenario (highest $\e(R)$ and lowest $\var(R)$) and worst-case scenario (lowest $\e(R)$ and highest $\var(R)$).

@@

ggplot(saturation_days, aes(n_days, hr)) +
  facet_wrap(~ case, nrow = 1) +
  geom_vline(xintercept = 100, color = 'darkorange') +
  geom_smooth(method = 'gam', color = 'black',
              formula = y ~ s(x, bs = 'cs', k = 10),
              method.args = list(family = Gamma(link = 'log'))) +
  geom_point(alpha = 0.3) +
  scale_x_continuous(expression(Number~of~days~sampled~(log[2]~scale)),
                     trans = 'log2', breaks = c(2, 16, 128, 1024),
                     limits = c(2, 1100)) +
  scale_y_continuous(expression(atop(Estimated~'space-use',
                                     requirements~(log[2]~scale))),
                     trans = 'log2')
```

```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated space-use requirements as a function of the number of days sampled for an animal in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an animal with the lowest $\\e(R)$ and highest $\\var(R)$ (right). In both cases, 100 days are sufficient to produce stable estimates of space-use requirements.", cache=FALSE, out.width='75%', fig.height=8, fig.width=12}
knitr::include_graphics('../figures/hr-over-days.png', rel_path = FALSE)
```

@@

This final section illustrates the location-scale GAM used to estimate the effects of resource abundance and stochasticity on the simulated space-use requirements. Although the model was fit to both the 95% and core (50%) home range estimates, the results were presented only for the 95% quantile for simplicity. The home range was caulculated using a Gaussian approximation using the simulated organism's positional variance:

$$\widehat{h_q} = -2 \log(1 - q) \pi \sigma^2_{pos},$$

where $widehat{h_q}$ is the estimated home range for the $q^{\text{th}}$ quantile and $\sigma^2_{pos}$ is the positional variance. Consequently, the estimated core home range estimate is simply a multiple of the 95% quantile:

@@

and therefore $\e(R)$ and $\var(R)$ would have similar effects (but about 4.322 times weaker) on the core home range than they do on the 95% home range. Note that we used this approximation because, unlike real organisms, the simulated organism did not use the available space selectively (e.g., spending more time near cell boundaries), so the Gaussian approximation is appropriate. When estimating space-use requirements based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix C on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].

@@

The sampling of the data is not uniform:

```{r, eval=TRUE, fig.height=4}
ggplot(sims, aes(mu, sigma2)) +
  geom_point(alpha = 0.3) +
  theme_bw() # Rmd only keeps theme_bw as default for a single plot
```

But it is not particularly problematic:

```{r, eval=TRUE}
ggplot(sims, aes(mu, sigma2)) +
  geom_hex(bins = 20) +
  scale_fill_viridis_c('Count', limits = c(0, NA), option = 'A',
                       direction = -1) +
  theme_bw()
```

The abundance of data for the average $\e(R)$ and $\var(R)$ did not create biases in the models, since removing such values had no appreciable effect on the model estimates (not shown). To estimate the effects of resource abundance and stochasticity on the simulated space-use requirements, we fit 3 location-scale GAMs (GAMLSs) with a location-scale gamma distribution (since home ranges are strictly positive) of increasing complexity. We hypothesized that the effect of $\e(R)$ would depend on $\var(R)$ (and the effect of $\var(R)$ would depend on $\e(R)$), since we expected organisms to respond to changes in $\mu(t)$ less in predictable environments than in stochastic ones. The three models are detailed below and compared using the Akaike Information Criterion and the Bayesian Information Criterion. Both information criteria demonstrate that the third model, which accounts for the interaction effects of $\e(R)$ and $\var(R)$, is the best model. Each of the three models are based on the directed acyclical graph (DAG) shown in figure B\@ref(fig:dag). Within a Bayesian framework, DAGs can be used to infer causality within statistical models [as the latter alone does not allow one to infer causality, see @mcelreath_statistical_2016]. The DAG we present in this appendix can be interpreted as follows: $\e(R)$ and $\var(R)$ are the controlled variables that determine both the amount of resources in a given encounter with resources, $R$, and, consequently, an organism's space-use requirements, $H$. Additionally, as indicated by the arrow going from $\e(R)$ to $\var(R)$, the effect of $\var(R)$ depends of the value of $\e(R)$ (and vice-versa, but arrows in DAGs have to be unidirectional).

@@

```{r, eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth'}
# marginal effects of mu and sigma only (linear on the log link scale)
m_1 <- gam(
  list(
    # predictor for the mean
    hr ~ quantile + mu + sigma2,
    # predictor for the scale (variance = scale * mean)
    ~ quantile + mu + sigma2),
  family = gammals(),
  data = sims,
  method = 'REML')

# marginal effects of mu and sigma only (nonlinear on the log link scale)
m_2 <- gam(
  list(
    hr ~ quantile + s(mu) + s(sigma2),
    ~ quantile + s(mu) + s(sigma2)),
  family = gammals(),
  data = sims,
  method = 'REML')

# marginal and interaction effects (nonlinear on the log link scale)
m_3 <- gam(
  list(
    hr ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5),
    ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5)),
  family = gammals(),
  data = sims,
  method = 'REML')

# check if CV = sigma/mu is better than sigma
m_4 <- gam(
  list(
    hr ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5),
    ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5)),
  family = gammals(),
  data = mutate(sims, cv = sqrt(sigma2) / mu),
  method = 'REML')

# ti term of the scale parameter looks over-fit and too uncertain
draw(m_4) & theme_bw() + theme(panel.grid = element_blank())
```

@@

```{r, eval=TRUE}
AIC(m_1, m_2, m_3, m_4) # not much of a difference between m_3 and m_4
BIC(m_1, m_2, m_3, m_4) # not much of a difference between m_3 and m_4
```

```{r, eval=TRUE}
# the difference in df (and significance) is in the ti.1() term
summary(m_3)
```

```{r, eval=TRUE}
summary(m_4)
```

```{r, eval=TRUE, fig.height=8, fig.width=8, out.width='65%'}
m <- m_3 # select the best model
```

@@

```{r gamls-terms, fig.cap = "Marginal and interaction effects of each of the terms present in the final GAMLS (on the log link scales). The dashed lines indicate the 95\\% Bayesian credible intervals. The data points are included as rug plots in all marginal terms and as points in the interaction terms.", eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth', echo=FALSE}
# using custom layout to make figure easier to read
layout(matrix(c(7, 1, 2, 3, 8, 4, 5, 6), nrow = 4, byrow = TRUE))
draw(m) & theme_bw() + theme(panel.grid = element_blank())
```

@@

```{r, eval=TRUE}
ggplot() +
  facet_grid(. ~ x, scales = 'free', switch = 'x',
             labeller = label_parsed) + # label below x axis
  geom_point(aes(value, hr), filter(sims_l, quantile == 'hr_95'),
             alpha = 0.1, color = pal[3]) +
  # 95% CIs are present but hidden under the estimates
  geom_ribbon(aes(value, ymin = hr_lwr, ymax = hr_upr, fill = x),
              filter(preds, quantile == 'hr_95'), alpha = 0.5) +
  geom_line(aes(value, hr, color = x),
            filter(preds, quantile == 'hr_95'), linewidth = 2) +
  scale_color_manual(values = pal) +
  scale_x_continuous(NULL, breaks = NULL) + 
  scale_y_continuous(bquote(paste(bold('Space-use requirements, '),
                                  bolditalic('H'))), breaks = NULL) +
  theme_bw() +
  # to make facet strip look like the x axis
  theme(legend.position = 'none',
        strip.background = element_blank(),
        strip.text = element_text(size = 12))
```

>>>>

---
title: \large How resource abundance and stochasticity affect organisms' range sizes
subtitle: "Appendix B: Simulations"
# author:
#   - name: Stefano Mezzini
#     email: stefano.mezzini@ubc.ca
#     institute: [biol, braes]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [biol, braes, cmps]
# institute:
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
# csl: 'the-american-naturalist.csl'
fontsize: 12pt
header-includes:
    - \renewcommand{\figurename}{Figure B\!\!} # for "Figure Bx."
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize} # smaller caption text
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    toc: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua' # for author affiliations
      - '--lua-filter=author-info-blocks.lua' # for author affiliations
---

@@

```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' home-range sizes were simulated. (a.) Tracks were simulated using an Integrated Ornstein-Uhlenbeck model (IOU model, an infinitely diffusive and continous-velocity movement model), starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square). (b.) Each time the track crossed into a new cell (green dots), the animal collected a random amount of resources that followed a Gamma distribution with common mean $\\mu (t)$ and variance $\\sigma^2 (t)$. (c.) Each time the animal collected more resources, its satiety (purple) increased. Once the animal collected sufficient resources, the animal stopped exploring (i.e., the track was truncated) and was allowed to return to $\\langle0, 0 \\rangle$ over the same amount of time needed to reach satiety. (D) The process was repeated 200 times (13 tracks pictured in this panel). (E) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h', cache=FALSE}
knitr::include_graphics('../figures/simulation-example.png', rel_path = FALSE)
```

@@

**Exploratory plot B\@ref(fig:dt-sensitivity)A.** The amount of time between encounters ranged from 1 second (the minimum sampling interval) to 24 minutes and 10 seconds. Approximately 93% of the encounters (500/536, excluding the first 3 events of each track) occurred with 30 or more seconds between events.

**Exploratory plot B\@ref(fig:dt-sensitivity)B.** Halving the sampling interval had little to no effect on the total number of encounters for $\Delta t \lessapprox 60 \,s.$

**Exploratory plot B\@ref(fig:dt-sensitivity)C.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks but large enough to avoid excessive amounts of data and an inflated amount of encounters when an animal was near cell boundaries.

**Exploratory plot B\@ref(fig:dt-sensitivity)D.** The three tracks used for these exploratory plots are sufficiently different that we considered them to be a representative sample of movement tracks simulated by the OUF model. All three tracks have a reasonable amount of both tortuous and directed movement.

@@

As the organism searched for resources, it encountered $R$ resources sampled from a family of Gamma distributions. The figure below shows the range in Gamma distributions for the minimum, median, and maximum mean (100, 150, 200) and variance ($2\,500, 126\,250, 250\,000$).

\renewcommand{\figurename}{Figure B\!\!\!\!}

```{r, eval=TRUE, echo=FALSE, fig.cap="Gamma distributions for the minimum, median, and maximum mean (100, 150, 200) and variance ($2\\,500, 126\\,250, 250\\,000$) with the corresponding coefficients of variation. The median and mode of the distributions increase with the mean but decrease with the variance. Greater coefficients of variation result in wider distributions with greater density near zero."}
setwd('..')
source('analysis/simulations/mean-variance-trends-panel-data.R')
source('functions/rgamma2.R')
setwd('writing')

expand_grid(mu = quantile(d55$mu, c(0, 0.5, 1)),
            sigma2 = quantile(d55$sigma2, c(0, 0.5, 1))) %>%
  mutate(sample = map2(mu, sigma2,
                       \(.m, .s) tibble(R = rgamma2(.m, .s, 1e5))),
         parameters = paste('list(', '\U03BC~"="~', mu, ', ',
                            'sigma^2~"="~', round(sigma2), ', ',
                            'sigma/\U03BC~"="~', round(sqrt(sigma2) / mu, 2),
                            ')') %>%
           factor(., levels = unique(.))) %>%
  unnest(sample) %>%
  ggplot(aes(R)) +
  facet_wrap(~ parameters, labeller = label_parsed, scales = 'fixed') +
  geom_density(fill = 'grey', bw = 5, na.rm = TRUE) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 399)) +
  scale_y_continuous('Density', expand = c(0, 0)) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

\renewcommand{\figurename}{Figure B\!\!}

@@

In this script, we check how many tracks are necessary to produce stable and accurate home-range size estimates. We do so by estimating the home ranges of animals for varying numbers of track in the best-case scenario (highest $\e(R)$ and lowest $\var(R)$) and worst-case scenario (lowest $\e(R)$ and highest $\var(R)$).

@@

ggplot(saturation_days, aes(n_days, hr)) +
  facet_wrap(~ case, nrow = 1) +
  geom_vline(xintercept = 100, color = 'darkorange') +
  geom_smooth(method = 'gam', color = 'black',
              formula = y ~ s(x, bs = 'cs', k = 10),
              method.args = list(family = Gamma(link = 'log'))) +
  geom_point(alpha = 0.3) +
  scale_x_continuous(expression(Number~of~days~sampled~(log[2]~scale)),
                     trans = 'log2', breaks = c(2, 16, 128, 1024),
                     limits = c(2, 1100)) +
  scale_y_continuous(expression(atop(Estimated~'home-range',
                                     size~(log[2]~scale))),
                     trans = 'log2')
```

```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated home-range sizes as a function of the number of days sampled for an animal in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an animal with the lowest $\\e(R)$ and highest $\\var(R)$ (right). In both cases, 100 days are sufficient to produce stable estimates of home-range size.", cache=FALSE, out.width='75%', fig.height=8, fig.width=12}
knitr::include_graphics('../figures/hr-over-days.png', rel_path = FALSE)
```

@@

This final section illustrates the location-scale GAM used to estimate the effects of resource abundance and stochasticity on the simulated home-range sizes. Although the model was fit to both the 95% and core (50%) home range estimates, the results were presented only for the 95% quantile for simplicity. The home range was caulculated using a Gaussian approximation using the simulated organism's positional variance:

$$\widehat{h_q} = -2 \log(1 - q) \pi \varsigma^2,$$

where $\widehat{h_q}$ is the estimated home range for the $q^{\text{th}}$ quantile and $\varsigma^2$ is the positional variance. Consequently, the estimated core home range estimate is simply a multiple of the 95% quantile:

@@

and therefore $\e(R)$ and $\var(R)$ would have similar effects (but about 4.322 times weaker) on the core home range than they do on the 95% home range. Note that we used this approximation because, unlike real organisms, the simulated organism did not use the available space selectively (e.g., spending more time near cell boundaries), so the Gaussian approximation is appropriate. When estimating home-range sizes based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix C on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].

@@

The sampling of the data is not uniform, but it is not particularly problematic.

```{r, eval=TRUE, fig.height=4, echo=FALSE}
ggplot(sims, aes(mu, sigma2)) +
  geom_point(alpha = 0.3) +
  theme_bw() # Rmd only keeps theme_bw as default for a single plot

ggplot(sims, aes(mu, sigma2)) +
  geom_hex(bins = 10) +
  scale_fill_viridis_c('Count', limits = c(0, NA), option = 'A',
                       direction = -1) +
  theme_bw()
```

The abundance of data for the average $\e(R)$ and $\var(R)$ did not create biases in the models, since removing such values had no appreciable effect on the model estimates (not shown). To estimate the effects of resource abundance and stochasticity on the simulated home-range sizes, we fit a location-scale GAMs (GAMLS) with a location-scale gamma distribution (since home ranges are strictly positive). We hypothesized that the effect of $\e(R)$ would depend on $\var(R)$ (and the effect of $\var(R)$ would depend on $\e(R)$), since we expected organisms to respond to changes in $\mu(t)$ less in predictable environments than in stochastic ones (fig. B\@ref(fig:dag)). Within a Bayesian framework, DAGs can be used to infer causality within statistical models [as the latter alone does not allow one to infer causality, see @mcelreath_statistical_2016]. The DAG we present in this appendix can be interpreted as follows: $\e(R)$ and $\var(R)$ are the controlled variables that determine both the amount of resources in a given encounter with resources, $R$, and, consequently, an organism's home-range sizes, $H$. Additionally, as indicated by the arrow going from $\e(R)$ to $\var(R)$, the effect of $\var(R)$ depends of the value of $\e(R)$ (and vice-versa, but arrows in DAGs have to be unidirectional).

@@

```{r, eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth'}
m <- gam(
  list(
    hr ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5),
    ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5)),
  family = gammals(),
  data = sims,
  method = 'REML')
```

```{r gamls-terms, fig.cap = "Marginal and interaction effects of each of the terms present in the final GAMLS (on the log link scales). The dashed lines indicate the 95\\% Bayesian credible intervals. The data points are included as rug plots in all marginal terms and as points in the interaction terms.", eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth', echo=FALSE}
draw(m) & theme_bw() + theme(panel.grid = element_blank())
```

@@

```{r, eval=TRUE}
summary(m)
```

@@

@@

```{r, eval=TRUE}
ggplot() +
  facet_grid(. ~ x, scales = 'free', switch = 'x',
             labeller = label_parsed) + # label below x axis
  geom_point(aes(value, hr), filter(sims_l, quantile == 'hr_95'),
             alpha = 0.1, color = pal[3]) +
  # 95% CIs are present but hidden under the estimates
  geom_ribbon(aes(value, ymin = hr_lwr, ymax = hr_upr, fill = x),
              filter(preds, quantile == 'hr_95'), alpha = 0.5) +
  geom_line(aes(value, hr, color = x),
            filter(preds, quantile == 'hr_95'), linewidth = 2) +
  scale_color_manual(values = pal) +
  scale_x_continuous(NULL, breaks = NULL) + 
  scale_y_continuous(bquote(paste(bold('Home-range size, '),
                                  bolditalic('H'))), breaks = NULL) +
  theme_bw() +
  # to make facet strip look like the x axis
  theme(legend.position = 'none',
        strip.background = element_blank(),
        strip.text = element_text(size = 12))
```
